---
title: "An introduction to SLOPE"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: SLOPE.bib
vignette: >
  %\VignetteIndexEntry{An introduction to SLOPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

The functions in this package solves problems of the type
$$
\mathrm{minimize}\left\{f(\beta) + J(\beta; \alpha,\lambda) \right\},
$$
where the second part of the objective is the
sorted L1-norm
$$
  J(\beta; \alpha,\lambda) =
    \alpha \sum_{j=1}^p \lambda_j \lvert \beta \rvert_{(j)},
$$
where $\alpha \in \mathbb{R}_{+}$, $\lambda \in \mathbb{R}_+^p$ and
$(j)$ represents an rank of the magnitudes of $\beta$ in descending order.
$\lambda$ controls the shape of the penalty sequence, which needs to be
non-increasing, and $\alpha$ controls the scale of that sequence.

Solving this problem is called SLOPE (Sorted L-One Penalized Estimation)
[@bogdan2015].

In this problem, $f(\beta)$ is a smooth and convex objective, which for this
package so far includes four models from the family of
generalized linear models:

* Gaussian regression,
* binomial regression,
* multinomial regression, and
* Poisson regression.

SLOPE is an extension of the lasso [@tibshirani1996] and has the ability
to lead to sparse solutions given a sufficiently strong regularization.
It is also easy to see that SLOPE reduces to the lasso if all elements
of the $\lambda$ vector are equal.

The lasso, however, has difficulties with correlated predictors [@jia2010]
but this is not the case with SLOPE, which handles this issue by clustering
predictors to the same magnitude. This effect is related to the
consecutive differences of the $\lambda$ vector: the larger the steps,
the more clustering behavior SLOPE exhibits.

## An example

In the following example, we will use the heart data set, for which the
response is a cardiac event. (The package contains several data sets
to exemplify modeling. Please see the examples in `SLOPE()`.) The main
function of the package is `SLOPE()`, which, more or less, serves as an
interface for code written in C++. There are many arguments in the
function and most of them relate to either the construction of the
regularization path or the penalty ($\lambda$) sequence used. Here we
will use the option `lambda = "bh"`, which used the BH method detailed
in @bogdan2015 to select the sequence. (Note that it is also possible
to manually insert a sequence.)

```{r}
library(SLOPE)

x <- heart$x
y <- heart$y

fit <- SLOPE(x, y, family = "binomial", lambda = "bh")
```

The default print method gives a summary of the regularization path but
it is usually more informative to study a plot of the path.

```{r, fig.cap = "Regularization path for a binomial regression model fit to the heart data set.", fig.width = 6, fig.height = 5}
plot(fit)
```

## Cross-validation

To determine the strength of regularization, it is almost always necessary to
tune the $\lambda$ sequence using resampling. This package features two
methods for this: a specification for use with the **caret** package via
the function `caretSLOPE()` or `trainSLOPE()`. While former facilitate
easier comparison of SLOPE against other methods as well as a multitude
of options for resampling and measuring performance, it does not
allow for sparse predictor matrices. We will give an example of `trainSLOPE()`
here.

```{r}
set.seed(924)

x <- bodyfat$x
y <- bodyfat$y

tune <- trainSLOPE(
  x,
  y,
  q = c(0.1, 0.2),
  number = 5,
  solver = "admm",
  repeats = 2
)
```

As before, the plot method offers the best summary.

```{r, fig.cap = "Model tuning results from Gaussian SLOPE on the bodyfat dataset.", fig.width = 5.5, fig.height = 3}
plot(tune, measure = "mae") # plot mean absolute error
```

Printing the resulting object will display the optimum values

```{r}
tune
```


## False discovery rate

Under assumptions of orthonormality, SLOPE has been shown to control
false discovery rate (FDR) of non-zero coefficients (feature weights)
in the model [@bogdan2015]. It is in many ways analogous to the
Benjamini--Hochberg procedure for multiple comparisons.

Let's set up a simple experiment to see how SLOPE controls the FDR.
We randomly generate data sets with various proportions of
true signals. Under this Gaussian design with independently and identically
distributed columns in $X$, SLOPE should asymptotically
control FDR at the level given by the shape parameter $q$, which
we set to 0.1 in this example.

```{r, fig.cap = "Control of false discovery rate using SLOPE.", fig.width = 4}
# proportion of real signals
q <- seq(0.05, 0.5, length.out = 20)
fdr <- double(length(q))
set.seed(1)

for (i in seq_along(q)) {
  n <- 1000
  p <- n / 2
  alpha <- 1
  problem <- SLOPE:::randomProblem(n, p, q = q[i], alpha = alpha)

  x <- problem$x
  y <- problem$y
  signals <- problem$nonzero

  fit <- SLOPE(x,
    y,
    lambda = "gaussian",
    solver = "admm",
    q = 0.1,
    alpha = alpha / sqrt(n)
  )

  selected_slope <- which(fit$nonzeros)
  V <- length(setdiff(selected_slope, signals))
  R <- length(selected_slope)
  fdr[i] <- V / R
}

library(ggplot2)

ggplot(mapping = aes(q, fdr)) +
  geom_hline(yintercept = 0.1, lty = 3) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(y = "False Discovery Rate")
```

SLOPE seems to control FDR at roughly the specified level.

# References
