---
title: "Models in SLOPE"
author: "Johan Larsson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: SLOPE.bib
vignette: >
  %\VignetteIndexEntry{Models in SLOPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

## Models

Sorted L-One Penalized Estimation (SLOPE) is the procedure
of minimizing objectives of the type
$$
  \mathrm{minimize}\left\{F(\beta_0, \beta; X, y) + J(\beta; \alpha,\lambda) \right\},
$$
where 
$$
  J(\beta; \alpha,\lambda) =
    \alpha \sum_{j=1}^p \lambda_j \lvert \beta \rvert_{(j)},
$$
with $\alpha \in \mathbb{R}_{+}$, $\lambda \in \mathbb{R}_+^p$ and
$(j)$ represents an rank of the magnitudes of $\beta$ in descending order.
$\lambda$ controls the shape of the penalty sequence, which needs to be
non-increasing, and $\alpha$ controls the scale of that sequence.

$X$ and $Y$ are the design matrix and response matrix, respectively, which
are of dimensions $n \times p$ and $n \times m$ respectively. Except
for multinomial logistic regression, $m = 1$.

We assume that $F$ takes the following form:
\[
  F(\beta_0, \beta) = \frac{1}{n} \sum_{i=1}^n f(\beta_0 + x_i^\intercal \beta, y_i),
\]
where $f$ is a smooth and convex loss function from the family of
generalized linear models (GLMs), $x_i$ is the $i$th row
of the design matrix $X$, and $y_i$ is the $i$th row of the response matrix
$Y$.

SLOPE currently supports four different models from the family of generalized
linear models (GLMs):

- Least-squares (Gaussian) regression,
- Logistic regression,
- Multinomial logistic regression, and
- Poisson regression.

Due to the way GLMs are formulated, each model is defined by three components:
a loss function \(f(\eta, y)\), a link function \(g(\mu)\), and an inverse link
function \(g^{-1}(\eta)\). Here, \(\eta\) is the linear predictor, and \(\mu\) is the 
mean of the response variable.


| Model       | \(f(\eta, y)\)                                                                                       | \(g(\mu)\)                                | \(g^{-1}(\eta)\)                                       |
|-------------|------------------------------------------------------------------------------------------------------|-------------------------------------------|--------------------------------------------------------|
| Gaussian    | \(\frac{1}{2}(y - \eta)^2\)                                                                          | \(\mu\)                                   | \(\eta\)                                               |
| Binomial    | \(\log(1 + e^\eta) - \eta y\)                                                                        | \(\log \left(\frac{\mu}{1 - \mu}\right)\) | \(\frac{e^\eta}{1 + e^\eta}\)                          |
| Poisson     | \(e^\eta - \eta y\)                                                                                  | \(\log(\mu)\)                             | \(e^\eta\)                                             |
| Multinomial | \(\sum_{k=1}^{m-1}\left( \log \left( 1 +  \sum_{j=1}^{m-1} e^{\eta_j}\right) - y_k \eta_k  \right)\) | \(\log\left(\frac{\mu}{1 - \mu}\right) \) | \(\frac{\exp(\eta)}{1 + \sum_{j=1}^{m-1} e^{\eta_j}}\) |

: Loss functions, link functions, and inverse link functions for generalized
  linear models in the SLOPE package. Note that in the case of multinomial
  logistic regression, the input is vector-valued, and we allow \(\log\) and
  \(\exp\) to be overloaded to apply element-wise in these cases.

## Gaussian Regression

Gaussian regression, also known as least-squares regression, is used for
continuous response variables, and takes the following form:
$$
f(\beta, \beta_0; X, y) =
  \frac{1}{2n} \sum_{i=1}^n \left(y_i - x_i^\intercal \beta - \beta_0\right)^2.
$$

You select it by setting `family = "gaussian"` in the `SLOPE()` function.
In the following example, we fit a Gaussian SLOPE model to the `bodyfat` data
set, which is included in the package:

```{r}
library(SLOPE)

fit_gaussian <- SLOPE(
  x = bodyfat$x,
  y = bodyfat$y,
  family = "gaussian"
)
```

We can plot the solution path:

```{r}
#| fig-width: 5
#| fig-height: 4
plot(fit_gaussian)
```

We can also print a summary of the fitted model:

```{r}
summary(fit_gaussian)
```

SLOPE also contains standard methods for returning coefficients and making predictions:

```{r}
coef(fit_gaussian)

y_hat <- predict(fit_gaussian, x = bodyfat$x[1:5, ], alpha = 0.01)

plot(bodyfat$y[1:5], y_hat)
```

## Logistic Regression

Logistic regression is used for binary classification problems, and
takes the following form: 
$$
f(\beta_0, \beta; X, y) =
  \frac{1}{n} \sum_{i=1}^n \left\{
    \log\left(1 + \exp\left(x_i^T \beta + \beta_0\right)\right) - y_i x_i^T \beta
  \right\},
$$
where $X_i$ is the $i$th row of the design matrix $X$.

You select it by setting `family = "binomial"` in the `SLOPE()` function.

<!-- ```{r} -->
<!-- fit_logistic <- SLOPE( -->
<!--   x = heart$x, -->
<!--   y = heart$y, -->
<!--   family = "binomial" -->
<!-- ) -->
<!-- ``` -->

## References
